The Two Dualities of Computation: Negative and Fractional Types

### 2012-03-11 rational.tex

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Negative and Fractional Types}
\label{sec:neg}

The structure of this section and the next are the same. The general outline
is as follows:\footnote{We refer the reader to Selinger's excellent survey
  article of monoidal
  categories~\cite{springerlink:10.1007/978-3-642-12821-94} for the precise
  definitions.}
\begin{itemize}
\item As established in our earlier work~\cite{rc2011,infeffects}, the
  underlying categorical semantics of our core reversible language
  {{langRev}} is based on \emph{two} distinct symmetric monoidal structures,
  one with {{+}} as the monoidal tensor, and one with {{*}} as the monoidal
  tensor.
\item We extend each underlying symmetric monoidal structure to a
  \emph{compact closed} structure by adding a dual for each object and two
  special morphims traditionally called {{eta}} and {{eps}}. 
\end{itemize}
After presenting the extension at the syntactic level, we discuss the
categorical semantics informally via the graphical language. The operational
semantics is presented in Sec.~\ref{sec:rat}.

As will be detailed in the remainder of this section, the compact closed
structure provides several properties of interest: (i) morphisms or wires are
allowed to run from right to left, (ii) the structure admits a trace that,
depending on the situation, can be used to implement various notions of loops
and
recursion~\cite{joyal1996traced,Hasegawa:2009:TMC:1552068.1552069,Hasegawa:1997:RCS:645893.671607},
(iii) the structure includes an isomorphism showing that the dual operator is
an involution, and (iv) the structure is equipped with objects representing
higher-order functions and a morphism \textit{eval} that applies these
functional objects. Interestingly each monoidal structure provides the same
ingredients, resulting in two dualities, two traces, two involutions, and two
notions of higher-order functions.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Additive Duality in {{langRev}} }

@@

\paragraph*{Negative Information Flow.} 
The intuitive way to read the diagrams is that they reverse the flow of
information. In previous {{langRev}} circuits, information was always flowing
from left to right. The presence of {{eps+}} operations cause information to
flow from right to left which can be reversed again using {{eta+}}.  As an
example, the diagram representing the isomorphism from the introduction is:
\begin{center}
  \includegraphics{diagrams/algebra1.pdf}
\end{center}

% \paragraph*{Time Traveling.}
% A time traveling intuition is also applicable here: if view the
% left-to-right direction of information flow as the canonical forward
% direction of information, then the action of {{eps+}} allows values,
% aka information particles, to flow backwards in time. Particles that
% flow backwards in time get to see and interact with the history of
% particles that they coexist with, i.e., values that they exist
% multiplicatively with. Hence the isomorphism {{(-b1)*b2<->-(b1*b2)}}
% (whose construction is shown in Section \ref{sec:neg-constructions}).

%\subsection{Compact Closure with Additive Inverses}
\paragraph*{Coherence}
The coherence condition for compact closed categories may be visualized as
follows~\cite{springerlink:10.1007/978-3-642-12821-94}:
\begin{center}
  \includegraphics{diagrams/coherence.pdf}
\end{center}
It is intuitively clear that if the interpretations of {{eta+}} and {{eps+}}
are to reverse the flow of execution, then this coherence condition is
satisfied.

@@

\begin{multicols}{2}
\begin{center}
  \includegraphics{diagrams/trace.pdf}
\end{center}

@@

\end{multicols}

@@

Multiplicative duality is added to {{langRev}} by adding types
{{1/b}}, values {{1/v}} and isomorphisms {{eta*}} and {{eps*}} with
the following type judgements.

@@

In the usual interpretation of {{b1*b2}} we have both a value of type {{b1}}
and a value of type {{b2}}. This interpretation is maintained in the presence
of fractionals. Hence an {{eta* : 1 <-> (1/b) * b}} is to be viewed as a
fission point for a value of type {{b}} and its multiplicative inverse
{{1/b}}. This naturally raises the question -- which value of type {{b}} is
produced? The isomorphism {{eta*}} does not produce any specific value of
type {{b}}, instead it produces a place holder for a value such that the
value of type {{1/b}} is always the dual of the chosen one.

@@

@@

The constructions for functions, involution, and trace are identical,
replacing {{+}} by {{*}} mutatis mutandis.

% \paragraph*{Unification.}
% A computational interpretation of this is that {{langRevEE}} acts like
% a logic programming system and {{eta*}} acts as the site of creation
% of a new typed logic variable of type {{b}} and its corresponding dual
% of type {{1/b}}. These two values are free to move independently
% through a circuit, but any operations on value affect what operations
% are possible on the other since they are both related by an underlying
% logic variable. Dually, {{eps*}} acts as a unification site for values
% flowing along both branches. If the values are not exactly duals, then
% the system fails to unify and hence the program is stuck. If the
% values are exactly duals, then unification succeeds and {{eta*}}
% returns {{1}}.

@@

% In a conventional language, higher-order functions capture scope and can be
% represented as first-class values. In {{langRevEE}}, we have two notions of
% functions both of which share similarities with conventional higher-order
% functions.

@@

% \item
% Multiplicative functions, of the form {{b1 -o* b2}}, on the other hand
% are indeed first-class values. However these functions can only be
% linearly used, i.e., they represent a constraint that can be satisfied
% at most once.
% %%  in one `computational world'. It may be possible that to
% %% repeatedly invoke multiplicative functions with an appropriate
% %% recursive type whose unfoldings allow for repeated unification.
% \end{itemize}

@@

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Negative and Fractional Types}

% The structure of this section and the next are the same.

This section introduces the syntax and graphical languages for
negative types and fractionals.The general outline is as follows:
\footnote{We refer the reader to Selinger's excellent survey
  article of monoidal
  categories~\cite{springerlink:10.1007/978-3-642-12821-94} for the precise
  definitions.}
\begin{itemize}
\item As established in our earlier work~\cite{rc2011,infeffects}, the
  underlying categorical semantics of our core reversible language
  {{langRev}} is based on \emph{two} distinct symmetric monoidal structures,
  one with {{+}} as the monoidal tensor, and one with {{*}} as the monoidal
  tensor.
\item We extend each underlying symmetric monoidal structure to a
  \emph{compact closed} structure by adding a dual for each object and two
  special morphims traditionally called {{eta}} and {{eps}}. 
\end{itemize}
After presenting the extension at the syntactic level, we discuss the
%% <<<<<<< local
%% categorical semantics informally via the graphical language. The
%% operational semantics is presented in Sec.~\ref{sec:rat}. As will be
%% detailed in this sections \ref{sec:neg} and \ref{sec:frac}, the
%% compact closed structure provides several properties of interest: (i)
%% morphisms or wires are allowed to run from right to left, (ii) the
%% structure admits a trace that, depending on the situation, can be used
%% to implement various notions of loops and
%% =======
categorical semantics informally via the graphical language. The operational
semantics is presented in Sec.~\ref{sec:rat}.

As will be detailed in the remainder of this section, the compact closed
structure provides several properties of interest: (i) morphisms or wires are
allowed to run from right to left, (ii) the structure admits a trace that,
depending on the situation, can be used to implement various notions of loops
and
%% >>>>>>> other
recursion~\cite{joyal1996traced,Hasegawa:2009:TMC:1552068.1552069,Hasegawa:1997:RCS:645893.671607},
(iii) the structure includes an isomorphism showing that the dual
operator is an involution, and (iv) the structure is equipped with
objects representing higher-order functions and a morphism
\textit{eval} that applies these functional objects. Interestingly
each monoidal structure provides the same ingredients, resulting in
two dualities, two traces, two involutions, and two notions of
higher-order functions.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Additive Duality in {{langRev}} }
\label{sec:neg}

@@

%% <<<<<<< local
%% Hence, the intuitive way to read the diagrams is that they reverse the
%% flow of information -- i.e. they capture a notion of \emph{negative
%%   information flow}. In normal {{langRev}} circuits, information flows
%% from left-to-right. The operator {{eps+}} causes information to flow
%% from right-to-left.  The interpretation of {{eta+}} is the opposite:
%% it turns a right-to-left flow of information in a circuit into a
%% left-to-right flow.  The diagram below is an example of using {{eta+}}
%% and {{eps+}} and corresponds to the additive {{a <-> a}} isomorphism
%% from the introduction. (Here the {{0}} wires have been shown for
%% completeness.)
%% =======
\paragraph*{Negative Information Flow.} 
The intuitive way to read the diagrams is that they reverse the flow of
information. In previous {{langRev}} circuits, information was always flowing
from left to right. The presence of {{eps+}} operations cause information to
flow from right to left which can be reversed again using {{eta+}}.  As an
example, the diagram representing the isomorphism from the introduction is:
%% >>>>>>> other
\begin{center}
\scalebox{1.2}{
  \includegraphics{diagrams/algebra1.pdf}
}
\end{center}

% \paragraph*{Time Traveling.}
%% <<<<<<< local
%% A \emph{time traveling} intuition is also applicable here.  In a
%% normal cicruit, as computational steps are taken values flow from left
%% to right, i.e. computational time progresses from left to right.  The
%% action of {{eps+}} causes values, aka information particles, to flow
%% from right to left i.e. backwards in computational time.  The
%% interesting thing about this interpretation is that particles that
%% travel backwards in time get to see and interact with the history of
%% particles that they coexist with. This gives us an intuitive
%% interpretation of the isomorphism {{(-b1)*b2<->-(b1*b2)}} (see
%% Sec. \ref{sec:specific-constructions}) where it would otherwise seem
%% that {{-b1}} and {{b2}} move in opposite directions. The backward flow
%% of the {{-b1}} allows it to `see the past' of {{b2}} and is thus
%% equivalent to both particles moving backward in time.
%% =======
% A time traveling intuition is also applicable here: if view the
% left-to-right direction of information flow as the canonical forward
% direction of information, then the action of {{eps+}} allows values,
% aka information particles, to flow backwards in time. Particles that
% flow backwards in time get to see and interact with the history of
% particles that they coexist with, i.e., values that they exist
% multiplicatively with. Hence the isomorphism {{(-b1)*b2<->-(b1*b2)}}
% (whose construction is shown in Section \ref{sec:neg-constructions}).
%% >>>>>>> other

%\subsection{Compact Closure with Additive Inverses}
%% <<<<<<< local
%% \emph{Coherence for Compact Closure with Additive Duality.}  The
%% appropriate coherence condition for compact closed categories,
%% expressed as the triangle axiom, is satisfied by {{eta+}}/{{eps+}} and
%% may be visualized (modulo {{0}} wires, swapping etc.) as explained by
%% Selinger \cite{springerlink:10.1007/978-3-642-12821-94} as the
%% operational equivalence:

%% =======
\paragraph*{Coherence}
The coherence condition for compact closed categories may be visualized as
follows~\cite{springerlink:10.1007/978-3-642-12821-94}:
%% >>>>>>> other
\begin{center}
  \includegraphics{diagrams/coherence.pdf}
\end{center}
It is intuitively clear that if the interpretations of {{eta+}} and {{eps+}}
are to reverse the flow of execution, then this coherence condition is
satisfied.

@@

%% <<<<<<< local
%% \emph{Non-termination.}  Compact closed categories contain {{trace}}
%% operations and this implies that {{eta+}}/{{eps+}} allow the
%% construction of a {{trace+}} operator (shown in Section
%% \ref{sec:monoidal-constructions}).  As shown in previous work
%% \cite{infeffects}, non-terminating computation can be constructed from
%%      {{trace+}} and recursive types.
%% =======
\begin{multicols}{2}
\begin{center}
  \includegraphics{diagrams/trace.pdf}
\end{center}

@@

\end{multicols}
%% >>>>>>> other

@@

%% <<<<<<< local
%% %% The section follows the outline of the previous one, first presenting
%% %% the new syntax, and discussing the graphical language and the
%% %% properites implied by the underlying categorical structure.

%% =======
%% >>>>>>> other
Multiplicative duality is added to {{langRev}} by adding types
{{1/b}}, values {{1/v}} and isomorphisms {{eta*}} and {{eps*}}.

@@

%% <<<<<<< local
%% The usual interpretation of {{b1*b2}} we have both a value of type
%% {{b1}} and a value of type {{b2}}. This interpretation is maintained
%% in the presence of fractionals. Hence an {{eta* : 1 <-> (1/b) * b}} is
%% to be viewed as a fission point for a value of type {{b}} and its
%% multiplicative inverse {{1/b}}. This naturally raises the question --
%% which value of type {{b}} is produced? The isomorphism {{eta*}} does
%% not produce any specific value of type {{b}}, instead it produces a
%% placeholder for an unspecified value and dually, {{1/b}} is inhabited
%% by the placeholder that indicates the absence of the unspecified
%% value.
%% =======
In the usual interpretation of {{b1*b2}} we have both a value of type {{b1}}
and a value of type {{b2}}. This interpretation is maintained in the presence
of fractionals. Hence an {{eta* : 1 <-> (1/b) * b}} is to be viewed as a
fission point for a value of type {{b}} and its multiplicative inverse
{{1/b}}. This naturally raises the question -- which value of type {{b}} is
produced? The isomorphism {{eta*}} does not produce any specific value of
type {{b}}, instead it produces a place holder for a value such that the
value of type {{1/b}} is always the dual of the chosen one.
%% >>>>>>> other

@@

%% <<<<<<< local
%% =======
%% >>>>>>> other

%% <<<<<<< local
This can be understood in terms of the \emph{logic programming} as
follows. Operator {{eta* : 1 <-> (1/b)*b}} acts as the site of
creation of a new typed logic variable and constructs from it a value
of type {{b}} and its dual of type {{1/b}}. These two values are free
to move independently through the circuit. Operations on one value
affect what operations are possible on the other since they are both
related by an underlying logic variable. Dually, {{eps*}} acts as a
unification site for values of type {{b}} and {{1/b}} flowing along
both branches. If the values are not unifiable duals, then the
specific program sequence fails. Like with logic programming, one can
has many worlds where some might fail, while others make
progress. Sec. \ref{sec:rat} has more details.  If the values are
exactly duals, then unification succeeds and {{eta*}} returns {{1}}.

% \paragraph*{Entanglement and Anti-Particles.}

%% An analogy inspired by quantum mechanics is also applicable here.  The
%% operator {{eta*}} is a site of fission that creates \emph{an entangled
%%   particle and anti-particle}. These particles can be thought to be in
%% a \emph{superposition} of states determined by their type. Both flow
%% in the same direction in time.  In each possible world that the
%% particle exists, it takes on a specific value inhibiting its type and
%% correspondingly its entangled anti-particle takes the dual of the
%% specific value. Since the particles are entangled, actions on one
%% (such as transformation by application of an isomorphism) affects the
%% other in much the same sense as \emph{action at a distance}. The
%% `{{a-o*b}}' functions mentioned in the introduction, are first-class
%% values and they correspond to these entagled pairs. Finally {{eps*}}
%% operations are sites where corresponding particle and anti-particles
%% annihilate each other. While appealing as an analogy, this description
%% does not imply (or preclude) any formal connection with Physics.

%% We can ask the function to be applied to a value by
%% unifying the anti-particle with a particle in a well-known state and
%% examining is pair. We shall use this point of view in the development
%% our SAT solver (see Sec. \ref{sec:sat-solver}).

%%  This also leads to the exciting point of view that
%% entangled such particle/anti-particle pairs, which are first class
%% values, act as functions in a physical first-class sense.

%% Further the fission site does not create a particle in any specific
%% state, but in a superposition of all possible states as dictated by
%% its type.

% Since such pairs are first-class entities, they correspond to

%% \emph{Coherence for Compact Closure with Multiplicative Duality.}  As
%% before, the triangle axiom which is the coherence condition on compact
%% closed categories is applicable and is expressed below in terms of
%% {{eta*}}/{{eps*}} and the multiplicative monoid.
%% =======
%% >>>>>>> other

@@

\emph{Annihilation.}  
With {{eta*}} and {{eps*}}, we can construct a {{trace*}} operation
which gives us the ability to find the fixpoint of a constraint.  If
the constraint has no satisfying values, there is no fixpoint possible
-- we use the term \emph{annihilation} to describe the corresponding
undefined state of the program.

%% <<<<<<< local
Annihilation and non-termination describe undefined program
execution. While non-termination characterizes an iteration that fails
to terminate, annihilation characterizes a constraint that has no
fixpoint. The SAT-solver of Sec. \ref{sec:sat-solver} works by
annihilating only the program states that do not match the boolean
satisfiability constraint.
%% =======
% \paragraph*{Unification.}
% A computational interpretation of this is that {{langRevEE}} acts like
% a logic programming system and {{eta*}} acts as the site of creation
% of a new typed logic variable of type {{b}} and its corresponding dual
% of type {{1/b}}. These two values are free to move independently
% through a circuit, but any operations on value affect what operations
% are possible on the other since they are both related by an underlying
% logic variable. Dually, {{eps*}} acts as a unification site for values
% flowing along both branches. If the values are not exactly duals, then
% the system fails to unify and hence the program is stuck. If the
% values are exactly duals, then unification succeeds and {{eta*}}
% returns {{1}}.

@@

% In a conventional language, higher-order functions capture scope and can be
% represented as first-class values. In {{langRevEE}}, we have two notions of
% functions both of which share similarities with conventional higher-order
% functions.
%% >>>>>>> other

@@

%% <<<<<<< local
The constructions for functions, involution, and trace are identical,
replacing {{+}} by {{*}}, \textit{mutatis mutandis}.

%% \begin{enumerate}
%% \item 
%% \emph{Trace}
%% =======
%% % \item
%% % Multiplicative functions, of the form {{b1 -o* b2}}, on the other hand
%% % are indeed first-class values. However these functions can only be
%% % linearly used, i.e., they represent a constraint that can be satisfied
%% % at most once.
%% % %%  in one `computational world'. It may be possible that to
%% % %% repeatedly invoke multiplicative functions with an appropriate
%% % %% recursive type whose unfoldings allow for repeated unification.
%% % \end{itemize}
%% >>>>>>> other

@@

\paragraph*{Zero Information.}
As with {{eta+}}, we argue that {{eta*}} constructs no new
information, instead it gives us the ability to explore as many worlds
as permitted by the type {{b}}. So situate ourselves in any specific
world, we have to supply additional information in the form of
unifying the particle or anti-particle with a particle representing
known information. Hence only by supplying information to a an
entangled pair can we read information from it -- and always exactly
as much information as we supplied.

ADD MATH.

\emph{Zero Information.}  The intuitive explanation that {{eta+}} and
hence {{eps+}} introduces no information effects is that {{-b+b}} as a
type whose value is observable only if you `pay' {{b}} amount of
information to the {{-b}} branch. When {{b}} amount of information is
supplied {{(-b+b)+b=b}} amount of information is returned to
you. (TODO. more needs to be said here.)

For this reason, our treatment of negative values in the context of
{{langRev}} is particularly simple. We will have a type $0$ and an
isomorphism {{0 <-> a + (-a)}} that when used in the left-to-right direction
allows us to create a value and a corresponding debt out of nothing. Both the
value and its negative counterpart can flow anywhere in the system. Because
information is preserved, a closed program (which does not have any
``dangling'' negative values) will eventually have to match the negative
value with some corresponding value, effectively using the isomorphism in the
right-to-left direction.
